A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-10-28 jacobian_calc_CMAME.tex
Revised/updated unclear language, cleaned up table, made accuracy plots easier to compare, added new table column type

@@

The methods were compared by running two sets of test problems where each method would solve a problem concurrently. Each test problem simulated a $\SI{4}{\meter}$ by $\SI{0.25}{\meter}$ by $\SI{0.25}{\meter}$ meter block of a material undergoing tension along the axis parallel to the long dimension of the block. The test problems were set up in \emph{Perdigm} as quasistatic equilibrium problems, where the displacement boundary conditions used to apply tension were applied gradually, in `load steps', and an equilibrium solution for each load step was achieved before applying the next load step. The purpose of applying the load in gradual steps is a widely used technique to ensure the Newton's method is always initialized near the actual solution and therefore likely to converge. The material properties of the model were meant to replicate a typical metal and had values of $\SI{1.515e4}{\mega\pascal}$ for bulk modulus, and $\SI{7.813e4}{\mega\pascal}$ for shear modulus\note[JTF]{Michael, are these numbers correct?}. The peridynamic horizon, a length scale that effectively sets the bandwidth of tangent-stiffness in the computation, used in the test problems was always set to three times the nominal discretization size (i.e. node spacing in the particle discretization scheme). This results in a minimum tangent-stiffness matrix bandwidth of 7 (for an optimal node numbering scheme).

@@

Once the quantities-of-interest (QoIs) for comparison were determined, the task was to find a way to take measurements that satisfied our goals. The goals of collecting accuracy and speed measurements were achieved by developing and implementing metrics within the simulation program used in the study. The metric used to measure the accuracy of a tangent-stiffness matrix was the Frobenius norm (analogous to an $l^2$-norm except defined on a matrix) of the element-wise difference between the tangent-stiffness matrix produced by the method being evaluated and the tangent-stiffness matrix produced by the AD based method (i.e., the \emph{exact} derivatives), given that both methods were set upon the same problem. The lower the value of this metric, the closer the tangent-stiffness of the method being evaluated (i.e, CS, AD, FD)  was to the AD Jacobian, the more accurate the method was. The expression for the accuracy metric is
%
\begin{equation} 
  D = \sqrt{\sum_i \sum_j(K^{AD}_{ij} - K^{M}_{ij})^2}
\end{equation}
%
Where $D$ is distance, $K^{AD}$ is the tangent-stiffness matrix produced by the AD based method, $M$ is replaced with either CS, FD, or CD depending on the method being compared.  The metric used to compare the speed of the different methods was the time in seconds required to calculate the tangent-stiffness matrix\note[JTF]{Verify the correctness of the this statement.  What was here previously was too wordy}. A final basis of comparison used in the study called \emph{computational efficiency} is based on the ratio of calculation time to the total number of tangent-stiffness matrix elements.

@@

\subsubsection{Data Reduction} 
The output produced by \emph{Perdigm} including the additional accuracy and speed measurements was redirected from the console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).  Directories and filenames were chosen so that data would be easy to sort by individual test run.  Extraneous information was manually deleted from copied versions of the output text files so that they would be regular enough to be easily read by scripts written in \emph{Python}. These scripts averaged accuracy or speed data for a tests corresponding to a particular peridynamic node density over all iterations for that run, and then plotted this averaged data  as a function of peridynamic node density as it varied from test to test in the series.  Similarly, for the parallel tests, the data was averaged and plotted in much the same way, but as a function of number of compute cores used to solve the problem rather than as a function peridynamic node density which was held constant. These data reduction and plotting scripts can be found on the corresponding author's website.

Since samples were not repeated, because there was not resources for it, confidence intervals were not calculated. However, if results of iterations within the same test problem were considered samples of the same thing, then a confidence interval could be calculated. This assumption was only made for the purpose of reporting average values on the plots. But, a cursory review of data showed no range overlap among the methods for accuracy or speed data from any test run of either series. \note[JTF]{Revise this paragraph.}

\section{Results and Discussion}
%
This section contains the average results from the study presented in tabular and graphic format.  Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) non-zero matrix elements, load step average calculation time and load step average accuracy. The units for accuracy are derived from the
expression used to calculate the TS matrix elements and the units used to define bulk material properties, mentioned in Section~\ref{tcs}.
%
\note[JTF]{Michael, work on cleaning this table up for legibility}
\begin{table}    
        \caption{Averaged Results, Each Test} \label{tab:results}   
    \begin{tabular}{c c c c }
      Cores & TS Elements & Calc. Time (s) & Accuracy Diff. ($\si{\mega\pascal}$)\\ 
    \multicolumn{2}{c}{} & CS, CD, FD, AD & CS, CD, FD \\ \hline 
    1 & $\num{1.99E6}$  & 3.5, 2.7, 1.7, 1.6 & 1.92E-10, 1.21E-4, .137 \\ \hline
	1 & 4.25E6  & 6.2, 4.9, 3.2, 2.9 & 2.28E-10, 9.94E-4, .148 \\ \hline
	1 & 7.79E6  & 11.2, 8.9, 5.7, 5.2 & 2.38E-10, 1.59E-4, .145\\ \hline
	1 & 1.54E7  & 26.7, 21.0, 13.4, 12.2 & 2.33E-10, 4.61E-4, .12 \\ \hline
	1 & 2.00E7  & 28.1, 22.2, 14.4, 13.1 & 2.76E-10, 1.05E-3, .145 \\ \hline
	1 & 3.14E7  & 47.6, 37.7, 24.2, 21.9 & 2.64E-10, 1.65E-3, .133 \\ \hline
	1 & 4.01E7  & 55.6, 44.1, 28.4, 25.9 & 3.03E-10, 1.92E-3, .148 \\ \hline
	1 & 8.39E7  & 138.9, 109.6, 70.2, 64.0 & 3.63E-10, 1.64E-3, .123 \\ \hline
	1 & 1.65E8  & 277.3, 218.1, 139.4, 126.5 & 3.26E-10, 2.18E-3, .128 \\ \hline
	32 & 1.67E10  & 336.1, 277.7, 200.6, 233.1 & 6.21E-10, 1.52E-2, .176 \\ \hline
	64 & 1.67E10  & 169.9, 140.7, 102.0, 119.7 & 6.20E-10, 1.50E-2, .177 \\ \hline
	96 & 1.67E10  & 114.7, 95.0, 69.1, 79.7 & 6.18E-10, 1.50E-2, .177 \\ \hline
	128 & 1.67E10  & 86.4, 71.8, 52.4, 58.8 & 6.16E-10, 1.47E-2, .177 \\ \hline
    \end{tabular}
\end{table}

\subsection{Speed Data}
Speed measurements were taken according to the methods mentioned in Section~\ref{tcs}. The averaged results of these speed measurements for the serial tests can be seen in Fig.~\ref{fig:serial_speed}.  Average time in seconds to compute a TS element is plotted as a function of the number of nonzero TS elements.  CS turned out to be the slowest of the four methods, taking the longest average time per iteration,  followed by CD, FD and AD in that order. Because the abscissa of Fig.~\ref{fig:serial_speed} is on a log scale, it bears mentioning that while visually it may appear that the trend indicated by the progression of data points on the plot is non-linear, the relationship would be closer to linear \note[JTF]{??? Revise}. This bodes well for the scalability of each of the methods by themselves in \emph{Peridigm}. 
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_speed.pgf}}
  \caption{Serial test series speed measurements.}
  \label{fig:serial_speed}
\end{figure}
%
For the parallel simulations, averaged speed results appear in Fig.~\ref{fig:multi_speed}. For this series, rather than varying the number of nonzero TS matrix elements the number of compute cores used to solve the problem is increased, while the number of nonzero TS matrix elements remained constant.  CS again was the slowest method, followed by central-difference, automatic-differentiation and then forward-difference.  There is a somewhat surprise reversal of FD and AD for these parallel simulations.  Figure~\ref{fig:multi_speed} indicates that when looking at average iteration time, each of the methods scale linearly with the number of compute cores used for the given conditions of this group of tests. Again, these results show that each of the methods as implemented in \emph{Peridigm} scale well when additional processors are used. 
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/multi_speed.pgf}}
  \caption{Multicore test series speed measurements.}
  \label{fig:multi_speed}
\end{figure}
%
A speculative explanation as to why FD and AD have reversed places in the speed tests with respect to the serial results is that the AD data-structures carry a list of partial derivatives of the functional evaluations along side the functional evaluations themselves only to be processed at the end of the tangent-stiffness calculation when the derivatives are requested.  It is likely that the communication cost associated with message passing these extra terms is what causes the AD method to be slower that FD.

\subsection{Accuracy Data}
Accuracy measurements where taken according to the methods mentioned in Section~\ref{tcs}\note[JTF]{Change L2 to $l^2$ in the figure label}. The averaged results of these accuracy measurements for the serial tests can be seen in Fig.~\ref{fig:serial_accuracy}. Averaged Frobenius norm (referred to as $l^2$-norm in the figure labels for conciseness) of the element-wise difference between a given TS matrix and the TS matrix produced by the AD method is plotted as a function of the number of nonzero TS matrix elements. CS was shown to be the most accurate of the methods when compared to AD, CS resulting in several orders of magnitude less difference than resulted from using the CD method and as expected FD method was the least accurate. This accuracy ranking order was maintained for each test in the series. It should be pointed out compared to the $l^2$-norms of any of the TS matrices by themselves, the magnitude of these differences for any of the methods is relatively small. These values appear in the simulation data which is available on the corresponding author's website. The effect of these accuracy differences on convergence rate of the Newton solver was not determined.  Additionally, we must point out that the accuracy of FD and CD could be improved by using a smaller probe distance, $h$; however, if $h$ is too small, then the methods can suffer from severe effects, including solution divergence, of the round-off error associated with subtracting two numbers that are very close to one another.  The value of $h$ used in this study is default value used in \emph{Peridigm} which is heuristically derived based on computational node spacing.  One of the beauties of the CS method is that there is no mystery as how to choose, $h$, it cannot be too small, therefore it can be set to machine epsilon and the analysis performed without the worry of issues associated with round-off error.
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_accuracy.pgf}}
  \caption{Serial test series accuracy measurements.}
  \label{fig:serial_accuracy}
\end{figure}
%
The averaged results of the accuracy measurements taken for the parallel test series can be seen in Fig.~\ref{fig:multi_accuracy}\note[JTF]{Change L2 to $l^2$ and make the ordinate axis the same scale as Figure 3}. The ordinate axis units remain unchanged from Fig.~\ref{fig:serial_accuracy}, but the absicca now indicates number of compute cores used to solve the test problem. Accuracy trends matched those for the single core tests. 
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/multi_accuracy.pgf}}
  \caption{Multicore test series accuracy measurements.}
  \label{fig:multi_accuracy}
\end{figure}

>>>>

\usepackage{array}

@@

The methods were compared by running two sets of test problems where each method would solve a problem concurrently. Each test problem simulated a $\SI{4}{\meter}$ by $\SI{0.25}{\meter}$ by $\SI{0.25}{\meter}$ meter block of a material undergoing tension along the axis parallel to the long dimension of the block. The test problems were set up in \emph{Perdigm} as quasistatic equilibrium problems, where the displacement boundary conditions used to apply tension were applied gradually, in `load steps', and an equilibrium solution for each load step was achieved before applying the next load step. The purpose of applying the load in gradual steps is a widely used technique to ensure the Newton's method is always initialized near the actual solution and therefore likely to converge. The fictional material represented in the model had values of $\SI{1.515e4}{\mega\pascal}$ for bulk modulus, and $\SI{7.813e4}{\mega\pascal}$ for shear modulus. The peridynamic horizon, a length scale that effectively sets the bandwidth of tangent-stiffness in the computation, used in the test problems was always set to three times the nominal discretization size (i.e. node spacing in the particle discretization scheme). This results in a minimum tangent-stiffness matrix bandwidth of 7 (for an optimal node numbering scheme).

@@

Once the quantities-of-interest (QoIs) for comparison were determined, the task was to find a way to take measurements that satisfied our goals. The goals of collecting accuracy and speed measurements were achieved by developing and implementing metrics within the simulation program used in the study. The metric used to measure the accuracy of a tangent-stiffness matrix was the Frobenius norm (analogous to an $l^2$-norm except defined on a matrix) of the element-wise difference between the tangent-stiffness matrix produced by the method being evaluated and the tangent-stiffness matrix produced by the AD based method (i.e., the \emph{exact} derivatives), given that both methods were set upon the same problem. The lower the value of this metric, the closer the tangent-stiffness of the method being evaluated (i.e, CS, AD, FD)  was to the AD Jacobian, the more accurate the method was. The expression for the accuracy metric is
%
\begin{equation} 
  D = \sqrt{\sum_i \sum_j(K^{AD}_{ij} - K^{M}_{ij})^2}
\end{equation}
%
Where $D$ is distance, $K^{AD}$ is the tangent-stiffness matrix produced by the AD based method, $M$ is replaced with either CS, FD, or CD depending on the method being compared.  The metric used to compare the speed of the different methods was the time in seconds required to calculate the tangent-stiffness matrix. A final basis of comparison used in the study called \emph{computational efficiency} is based on the ratio of calculation time to the total number of tangent-stiffness matrix elements.

@@

\subsubsection{Data Reduction} 
The output produced by \emph{Perdigm} including the additional accuracy and speed measurements was redirected from the console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).  Directories and filenames were chosen so that data would be easy to sort by individual test run.  Extraneous information was manually deleted from copied versions of the output text files so that they would be regular enough to be easily read by scripts written in \emph{Python}. These scripts averaged accuracy or speed data for a tests corresponding to a particular peridynamic node density over all iterations for that run, and then plotted this averaged data  as a function of peridynamic node density as it varied from test to test in the series.  Similarly, for the parallel tests, the data was averaged and plotted in much the same way, but as a function of number of compute cores used to solve the problem rather than as a function peridynamic node density which was held constant. Confidence intervals were not calculated on the measurements. The main reason is that network traffic from other users makes running tests on the HPC cluster a time varying process. These data reduction and plotting scripts can be found on the corresponding author's website.

\section{Results and Discussion}
%
This section contains the average results from the study presented in tabular and graphic format.  Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) non-zero matrix elements, load step average calculation time and load step average accuracy. The units for accuracy are derived from the
expression used to calculate the TS matrix elements and the units used to define bulk material properties, mentioned in Section~\ref{tcs}.
%
\note[JTF]{Michael, work on cleaning this table up for legibility}
\note[Michael]{Columns are now separated out and can be fine tuned}

\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\begin{table}    
        \caption{Averaged Results, Each Test} \label{tab:results}   
       \hskip-2.0cm\begin{tabular}{C{1.5cm} C{1.88cm} C{1.5cm} C{1.5cm} C{1.5cm} C{1.5cm} C{1.88cm} C{1.88cm} C{1.5cm}}
            Cores & TS Elements & Calc. Time (s) & & & & Accuracy Diff. ($\si{\mega\pascal}$) & & \\ 
        \multicolumn{2}{l}{} & CS & CD & FD & AD & CS & CD & FD \\ \hline 
        1 & $\num{1.99E6}$  & 3.5 & 2.7 & 1.7 & 1.6 & $\num{1.92E-10}$ & $\num{1.21E-4}$ & .137 \\ \hline
        1 & $\num{4.25E6}$  & 6.2& 4.9& 3.2& 2.9 & $\num{2.28E-10}$& $\num{9.94E-4}$ & .148 \\ \hline
        1 & $\num{7.79E6}$  & 11.2& 8.9& 5.7& 5.2 & $\num{2.38E-10}$& $\num{1.59E-4}$ & .145\\ \hline
        1 & $\num{1.54E7}$  & 26.7& 21.0& 13.4& 12.2 & $\num{2.33E-10}$ & $\num{4.61E-4}$ & .12 \\ \hline
        1 & $\num{2.00E7}$  & 28.1& 22.2& 14.4& 13.1 & $\num{2.76E-10}$ & $\num{1.05E-3}$ & .145 \\ \hline
        1 & $\num{3.14E7}$  & 47.6& 37.7& 24.2& 21.9 & $\num{2.64E-10}$ & $\num{1.65E-3}$ & .133 \\ \hline
        1 & $\num{4.01E7}$  & 55.6& 44.1& 28.4& 25.9 & $\num{3.03E-10}$ & $\num{1.92E-3}$ & .148 \\ \hline
        1 & $\num{8.39E7}$  & 138.9& 109.6& 70.2& 64.0 & $\num{3.63E-10}$ & $\num{1.64E-3}$ & .123 \\ \hline
        1 & $\num{1.65E8}$  & 277.3& 218.1& 139.4& 126.5 & $\num{3.26E-10}$ & $\num{2.18E-3}$ & .128 \\ \hline
        32 & $\num{1.67E10}$  & 336.1& 277.7& 200.6& 233.1 & $\num{6.21E-10}$ & $\num{1.52E-2}$ & .176 \\ \hline
        64 & $\num{1.67E10}$  & 169.9& 140.7& 102.0& 119.7 & $\num{6.20E-10}$ & $\num{1.50E-2}$ & .177 \\ \hline
        96 & $\num{1.67E10}$  & 114.7& 95.0& 69.1 & 79.7 & $\num{6.18E-10}$ & $\num{1.50E-2}$ & .177 \\ \hline
        128 & $\num{1.67E10}$  & 86.4& 71.8& 52.4 &58.8 & $\num{6.16E-10}$ & $\num{1.47E-2}$ & .177 \\ \hline
    \end{tabular}
\end{table}

\subsection{Speed Data}
Speed measurements were taken according to the methods mentioned in Section~\ref{tcs}. The averaged results of these speed measurements for the serial tests can be seen in Fig.~\ref{fig:serial_speed}.  Average time in seconds to compute a TS element is plotted as a function of the number of nonzero TS elements.  CS turned out to be the slowest of the four methods, taking the longest average time per iteration,  followed by CD, FD and AD in that order. Calculation time appears to have a linear relationship with problem size, judging by the appearance of the plot shown in Fig.~\ref{fig:serial_speed}. This bodes well for the scalability of each of the methods by themselves in \emph{Peridigm}. 
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_speed.pgf}}
  \caption{Serial test series speed measurements.}
  \label{fig:serial_speed}
\end{figure}
%
For the parallel simulations, averaged speed results appear in Fig.~\ref{fig:multi_speed}. For this series, rather than varying the number of nonzero TS matrix elements the number of compute cores used to solve the problem is increased, while the number of nonzero TS matrix elements remained constant.  CS again was the slowest method, followed by central-difference, automatic-differentiation and then forward-difference.  There is a somewhat surprise reversal of FD and AD for these parallel simulations.  Figure~\ref{fig:multi_speed} indicates that when looking at average iteration time, each of the methods scale linearly with the number of compute cores used for the given conditions of this group of tests. Again, these results show that each of the methods as implemented in \emph{Peridigm} scale well when additional processors are used. 
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/multi_speed.pgf}}
  \caption{Multicore test series speed measurements.}
  \label{fig:multi_speed}
\end{figure}
%
A speculative explanation as to why FD and AD have reversed places in the speed tests with respect to the serial results is that the AD data-structures carry a list of partial derivatives of the functional evaluations along side the functional evaluations themselves only to be processed at the end of the tangent-stiffness calculation when the derivatives are requested.  It is likely that the communication cost associated with message passing these extra terms is what causes the AD method to be slower that FD.

\subsection{Accuracy Data}
Accuracy measurements where taken according to the methods mentioned in Section~\ref{tcs}. The averaged results of these accuracy measurements for the serial tests can be seen in Fig.~\ref{fig:serial_accuracy}. Averaged Frobenius norm (referred to as $l^2$-norm in the figure labels for conciseness) of the element-wise difference between a given TS matrix and the TS matrix produced by the AD method is plotted as a function of the number of nonzero TS matrix elements. CS was shown to be the most accurate of the methods when compared to AD, CS resulting in several orders of magnitude less difference than resulted from using the CD method and as expected FD method was the least accurate. This accuracy ranking order was maintained for each test in the series. It should be pointed out compared to the $l^2$-norms of any of the TS matrices by themselves, the magnitude of these differences for any of the methods is relatively small. These values appear in the simulation data which is available on the corresponding author's website. The effect of these accuracy differences on convergence rate of the Newton solver was not determined.  Additionally, we must point out that the accuracy of FD and CD could be improved by using a smaller probe distance, $h$; however, if $h$ is too small, then the methods can suffer from severe effects, including solution divergence, of the round-off error associated with subtracting two numbers that are very close to one another.  The value of $h$ used in this study is default value used in \emph{Peridigm} which is heuristically derived based on computational node spacing.  One of the beauties of the CS method is that there is no mystery as how to choose, $h$, it cannot be too small, therefore it can be set to machine epsilon and the analysis performed without the worry of issues associated with round-off error.
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_accuracy.pgf}}
  \caption{Serial test series accuracy measurements.}
  \label{fig:serial_accuracy}
\end{figure}
%
The averaged results of the accuracy measurements taken for the parallel test series can be seen in Fig.~\ref{fig:multi_accuracy}. The ordinate axis units remain unchanged from Fig.~\ref{fig:serial_accuracy}, but the absicca now indicates number of compute cores used to solve the test problem. Accuracy trends matched those for the single core tests. 
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/multi_accuracy.pgf}}
  \caption{Multicore test series accuracy measurements.}
  \label{fig:multi_accuracy}
\end{figure}
